---
title: "DiagrammeR and economiccomplexity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Graphviz and mermaid in DiagrammeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

### Bipartite network projections

This example is a simplified version, extended to DiagrammeR, of the examples
from pacha.dev/economiccomplexity. The data used here corresponds to 
World Trade and Per-Capita GDP averages between 1998 and 2000 included in
`economiccomplexity::`.

```{r create-network}
library(DiagrammeR)
library(economiccomplexity)
library(igraph)

# Step 1: Obtain Balassa Index
bi <- balassa_index(world_trade_avg_1998_to_2000)

# Step 2: Obtain proximity matrices
pro <- proximity(bi)

# Step 3: Create bipartite network projections
net <- projections(pro$proximity_country, pro$proximity_product)
E(net$network_country)[1:5]
E(net$network_product)[1:5]
```

### Data aggregation for edge size

```{r alter-edge-size-1, fig.width=7, fig.height=7}
set.seed(200100)

aggregated_countries <- aggregate(
  world_trade_avg_1998_to_2000$value,
  by = list(country = world_trade_avg_1998_to_2000$country),
  FUN = sum
)

aggregated_countries <- setNames(aggregated_countries$x,
                                 aggregated_countries$country)

V(net$network_country)$size <- aggregated_countries[
  match(V(net$network_country)$name, names(aggregated_countries))]
```

```{r alter-edge-size-2, fig.width=10, fig.height=10}
aggregated_products <- aggregate(
  world_trade_avg_1998_to_2000$value,
  by = list(country = world_trade_avg_1998_to_2000$product),
  FUN = sum
)

aggregated_products <- setNames(aggregated_products$x,
                                aggregated_products$country)

V(net$network_product)$size <- aggregated_products[
  match(V(net$network_product)$name, names(aggregated_products))]
```

### Using Diagrammer to visualize

The last section returned a list containing two igraph objects. These
are almost ready to create the visualization.

```{r visualize-with-diagrammer-1, fig.width=10, fig.height=10}
# Step 1: Remove names and add labels (or it won't render well)
V(net$network_country)$label = V(net$network_country)$name
V(net$network_country)$name = factor(V(net$network_country)$name,
  levels = as.character(V(net$network_country)$name))

# Step 2: Visualize
vis_country <- from_igraph(net$network_country) %>% 
  set_node_attrs(
    node_attr = fillcolor,
    values = "SteelBlue")

# get_node_attrs(vis_country, "size") ok

vis_country %>% 
  render_graph(
    layout = "kk",
    title = "Proximity Based Network Projection for Countries"
  )
```

```{r visualize-with-diagrammer-2, fig.width=10, fig.height=10}
# Step 1: Remove names and add labels (or it won't render well)
V(net$network_product)$label = V(net$network_product)$name
V(net$network_product)$name = factor(V(net$network_product)$name,
  levels = as.character(V(net$network_product)$name))

# Step 2: Visualize
vis_product <- from_igraph(net$network_product) %>% 
  set_node_attrs(
    node_attr = fillcolor,
    values = "Crimson")

# get_node_attrs(vis_product, "size") ok

vis_product %>% 
  render_graph(
    layout = "kk",
    title = "Proximity Based Network Projection for Countries"
  )
```
